Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-01-26 chapters/chapter06.tex
HMMs explained

%TODO: explain hidden markov models
\begin{remark}{Infinite hidden Markov models}
Hidden Markov model (HMM) defines a probability distribution over sequences of observations

\end{remark}

>>>>

\begin{remark}{Infinite hidden Markov models}
Hidden Markov model (HMM) $\left\{ X\left(t\right):\; t=1,2,\ldots\right\}$ defines a probability distribution over sequences of observations $\mathbf{X}=\left\{ X(1), \ldots,X(t) \right\}$ by invoking another sequence of unobserved (\emph{hidden}) discrete state variables $\left\{ C\left(t\right):\; t=1,2,\ldots\right\}$.

% Let us denote the past states of these processes  from time $1$ to time $t$ by 
The model is defined by three parts: first the \emph{transition matrix} of the hidden process, which satisfies the Markov property, i.e.:

\begin{align}
% \mathbf{P}=\left\{ p_{ij}\right\} ,\text{where }p_{ij}=&P\left\{ C\left(t\right)|C\left(t-1\right),\ldots,C\left(1\right)\right\} \nonumber \\ 
% =&P\left\{ C\left(t\right)|C\left(t-1\right)\right\} , 
\mathbf{P}=\left\{ p_{ij}\right\} ,\text{where }p_{ij}=& P\left\{ C\left(t\right)=j|C\left(t-1\right)=i,C\left(t-2\right),\ldots \right\} \nonumber \\ 
=& P\left\{ C\left(t\right)=j|C\left(t-1\right)=i\right\} 
\label{eq:hmm_trans}
\end{align}

\noindent
secondly the \emph{emitting matrix} of the `state-dependent process', which depends only on the current state of the hidden process and not on the previous states, i.e.:

\begin{align}
% \mathbf{X}=\left\{ x_{ij}\right\} ,\text{where }x_{ij}=& P\left\{ X\left(t\right)|C\left(t\right),X\left(t-1\right),\ldots,X\left(1\right)\right\} \nonumber \\
% =&P\left\{ X\left(t\right)|C\left(t\right)\right\} ,
\mathbf{X}=\left\{ x_{jk}\right\} ,\text{where }x_{jk}=& P\left\{ X\left(t\right)=k|C\left(t\right)=j,X\left(t\right),X\left(t-1\right),\ldots \right\} \nonumber \\
=&P\left\{ X\left(t\right)=k|C\left(t\right)=j\right\} ,
\label{eq:hmm_emit}
\end{align}

\noindent
and lastly the vector of starting probabilities of being in some state $i$ at the beginning, i.e.:

\begin{align}
\Pi=\left\{ \pi_{i}\right\} ,\text{where }\pi_{i}=P\left\{ C\left(t=1\right)=i\right\} 
\label{eq:hmm_start}
\end{align}

\noindent
If the Markov chain $\left\{ C\left(t\right) \right\}$ has $m$ states we call $\left\{ X\left(t\right) \right\}$ an $m$-state HMM and $\lambda=\left(\mathbf{P},\mathbf{X},\Pi\right)$ is typically used as a compact notation for an HMM.
Extending  the \emph{hidden} $\left\{ C\left(t\right) \right\}$ process to a countably infinite number of hidden states for leads to the infinite HMM's.
\end{remark}
